---
title: "p8105_hw2_sm5975"
author: "Sania Sheifuddin"
date: "2026-10-07"
output: github_document
---

Read and clean the data; retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert the entry variable from character (YES vs NO) to a logical variable (case_match function may be useful).

Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far, and give the dimension (rows x columns) of the resulting dataset. Are these data tidy?

# Question 1 
```{r}
library(tidyverse)
nyctransit_df = 
    read_csv(
      "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", 
      na = c("NA", ".", ""), 
col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c"))

nyctransit_df =
  nyctransit_df |>
  janitor::clean_names() |>
  select(
    line, station_name, station_latitude, station_longitude,
    route1:route11, entry, vending, entrance_type, ada
  ) |>
  mutate(entry = case_match(entry, "YES" ~ TRUE, "NO" ~ FALSE))
      
```
This dataset consists of information on each entrance and exit of NYC subway situations. Variables included are the following:  station name, station latitude/longitude, routes served, subway line, whether the entrance allows entry, if there is vending, the entrance types and ADA compliance. 

To clean the data further I standardized the variable names to only have the variables of interest. I also converted entry from a character variable to a logical variable. After I ran the chunk I observed that the final dataset has `r nrow(nyctransit_df)` rows and `r ncol(nyctransit_df)` columns. Since route information is from route 1 through route 11 the route number is stored in the column names. 

```{r}
nyctransit_df |>
  distinct(line, station_name) |>
  nrow()
```
From the dataset we can see that, there are `r nrow(distinct(nyctransit_df, line, station_name))` distinct stations. 

```{r}
nyctransit_df |>
  filter(ada) |>
  distinct(line, station_name) |>
  nrow()
```

From the dataset we can see that there are `r nrow(distinct(filter(nyctransit_df, ada), line, station_name))` ADA compliant stations.

```{r}
nyctransit_df |>
  filter(vending == "NO") |>
  pull(entry) |>
  mean()
```
The proportion of entrances/exits without vending that allow entry is `r round(mean(filter(nyctransit_df, vending == "NO")$entry), 3)`.

```{r}
#Reformatting data so that route number and route name are distinct variables
nyctransit_tidy_df =nyctransit_df |>
  mutate(across(route8:route11, as.character)) |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name",
    names_prefix = "route",
    values_drop_na = TRUE
  )

# Distinct stations that serve the A train
nyctransit_tidy_df |>
  filter(route_name == "A") |>
  distinct(station_name, line) |>
  nrow()

# A train stations that are ADA compliant
nyctransit_tidy_df |>
  filter(route_name == "A", ada) |>
  distinct(station_name, line) |>
  nrow()
```

From the dataset provided we can see that, `r nyctransit_tidy_df |> filter(route_name == "A") |> distinct(station_name, line) |> nrow()` distinct stations serve the A train.

Of the stations that serve the A train, `r nyctransit_tidy_df |> filter(route_name == "A", ada) |> distinct(station_name, line) |> nrow()` are ADA compliant.

# Question 2

```{r}
#Read and Clean Mr. Trash Wheel sheet

library(readxl)
mrtrash_wheel_df =
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N742"
  ) |>
  janitor::clean_names() |> 
 drop_na(dumpster) |>
  mutate(sports_balls = as.integer(round(sports_balls)))
```
```{r}

#Combining with Mr. Trash Wheel Dataset to produce a single tidy dataset
library(tidyverse)
library(readxl)

mrtrash_wheel_df = 
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N742"
  ) |>
  janitor::clean_names() |> 
 drop_na(dumpster) |>
  mutate(
    sports_balls = as.integer(round(sports_balls)),
         year = as.integer (year),
         month = str_trim(month),
         trash_wheel = "Mr. Trash Wheel"
  )

proftrash_wheel_df =
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M139",
    col_types = c(
      "numeric", "text", "numeric", "date", "numeric", "numeric",
      "numeric", "numeric", "numeric", "numeric", "numeric", "numeric",
      "numeric"
    )
  ) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.integer(year),
    month = str_trim(month),
    trash_wheel = "Professor Trash Wheel"
  )

gwynndatr_df = 
  read_excel(
     "data/202610 Trash Wheel Collection Data.xlsx",
     sheet = "Gwynnda the Good Wheel of the W",
     range = "A2:L394"
  ) |> 
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(
    year = as.integer(year),
    month = str_trim(month),
    trash_wheel = "Gwynnda"
  )

trash_wheel_df = 
  bind_rows (mrtrash_wheel_df, proftrash_wheel_df, gwynndatr_df) |> 
  relocate (trash_wheel)

```
```{r}
 # the total weight of trash collected by Professor Trash Wheel
prof_totalweight =
  trash_wheel_df |>
  filter(trash_wheel == "Professor Trash Wheel") |>
  pull(weight_tons) |>
  sum(na.rm = TRUE)

#the total number of cigarette butts collected by Gwynnda in June of 2022
gwynnda_2022cigg =
  trash_wheel_df |>
  filter(trash_wheel == "Gwynnda", month == "June", year == 2022) |>
  pull(cigarette_butts) |>
  sum(na.rm = TRUE)

```
The final combined dataset, `trash_wheel_df`, contains `r nrow(trash_wheel_df)` observations, where each observation is a single dumpster collection by Mr. Trash Wheel, Professor Trash Wheel, or Gwynnda, with `r ncol(trash_wheel_df)` variables. The data span `r year(min(trash_wheel_df$date))` to `r year(max(trash_wheel_df$date))`. Key variables include `trash_wheel` (which wheel collected the trash), `dumpster` (the dumpster number), `month`, `year`, and `date`, along with `weight_tons` and `volume_cubic_yards`, and counts of specific items such as `plastic_bottles`, `cigarette_butts`, `wrappers`, and `homes_powered`. The variable `sports_balls` is available only for Mr. Trash Wheel, and `glass_bottles` is not recorded for Gwynnda, so those values are `NA` in the combined data. For available data, Professor Trash Wheel collected a total of `r round(prof_totalweight, 2)` tons of trash. Gwynnda collected a total of `r format(gwynnda_2022cigg, big.mark = ",")` cigarette butts in June of 2022.

#Question 3 

```{r}
mci_df =
  read_csv(
    "data/data_mci/MCI_baseline.csv",
    skip = 1,
    na = "."
  ) |>
  janitor::clean_names() |>
  mutate(
    sex = replace(sex, sex == 0, "Female"),
    sex = replace(sex, sex == 1, "Male"),
    sex = factor(sex),
    apoe4 = replace(apoe4, apoe4 == 0, "Non-carrier"),
    apoe4 = replace(apoe4, apoe4 == 1, "Carrier"),
    apoe4 = factor(apoe4))

mcibase_df =
  mci_df |>
  filter(is.na(age_at_onset) | age_at_onset > current_age)

#  Number of participants recruited, and of these how many develop MCI
n_recruit = nrow(mci_df)
n_eligible = nrow(mcibase_df)
n_develop_mci = sum(!is.na(mcibase_df$age_at_onset))

# Average Baseline age 
mean_baseline_age = mean(mcibase_df$current_age)

#Proportion of women in the study that are APOE4 carrier
prop_women_apoe4 =
  mcibase_df |>
  filter(sex == "Female") |>
  summarize(prop = mean(apoe4 == "Carrier")) |>
  pull(prop)
```
I did `skip = 1` as the baseline dataset required to bypass a row of variable descriptions, and `na = "."` to point out missing ages at MCI onset which are coded as periods. Sex and APOE4 carrier status were recoded to labeled factors. `r n_recruit` participants were recruited. Participants whose age at MCI onset was at or before their baseline age already had MCI at baseline and were hence removed, leaving `r n_eligible` eligible participants. Of these, `r n_develop_mci` developed MCI during follow-up. The average baseline age was `r round(mean_baseline_age, 2)` years, and `r round(prop_women_apoe4, 3)` of women in the study were APOE4 carriers.

```{r}

mciamyloid_df = 
  read_csv(
    "data/data_mci/MCI_amyloid.csv",
    skip = 1,
    na = c("NA", "Na")) |> 
    janitor::clean_names() |> 
    rename (id= study_id) |> 
  pivot_longer(
    baseline:time_8,
    names_to = "time",
    names_prefix = "time_",
    values_to = "amyloid_ratio") |> 
  mutate(
    time =replace (time, time =="baseline","0"),
    time = as.numeric(time))
print(mciamyloid_df, n = 12)

```
The first row of biomarker file contains a description of the time variables, so it was skipped with `skip = 1` to make it more concise, and this consequently makes the second row the header. The missing biomarker values appear under both, "NA" and "Na", so both are technically missing. If not for that, the amyloid columns would then be read as character. Variable names were standardized with `clean_names()`, and the `study_id` was renamed `id`as that was what it was reffered to on dataset. The data were originally wide, with one column per visit (baseline, then 2, 4, 6, and 8 years after baseline), so they were turned into long format with a `time` variable and an `amyloid_ratio` variable. The resulting dataset has `r n_distinct(mciamyloid_df$id)` participants and `r nrow(mciamyloid_df)` rows, one per participant per visit. Many measurements are missing (`r sum(is.na(mciamyloid_df$amyloid_ratio))`), and participants do not all have the same visits, so follow-up is unbalanced.

```{r}
# Participants who appear only in baseline data 
anti_join(mcibase_df, mciamyloid_df, by = "id") |>
  distinct(id)

# Participants who appear only in amyloid data 
anti_join(mciamyloid_df, mcibase_df, by = "id") |>
  distinct(id)
```

Not everyone appear in both datasets. `r nrow(distinct(anti_join(mcibase_df, mciamyloid_df, by = "id"), id))` participants have baseline demographics but no biomarker data, and `r nrow(distinct(anti_join(mciamyloid_df, mcibase_df, by = "id"), id))` participants (IDs 484 to 495) have biomarker data but no baseline demographics. The amyloid-only participants cannot be checked against the inclusion criteria, since their age, sex, APOE4 status, and MCI history are unknown. This suggests the two files came from slightly different extracts of the study. Participants missing from either file cannot be used in analyses that need both demographics and biomarkers, so they are excluded when the datasets are combined.