---
title: "p8105_hw2_sm5975"
author: "Sania Sheifuddin"
date: "2026-10-07"
output: github_document
---

Read and clean the data; retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert the entry variable from character (YES vs NO) to a logical variable (case_match function may be useful).

Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far, and give the dimension (rows x columns) of the resulting dataset. Are these data tidy?

# Question 1 
```{r}
library(tidyverse)
nyctransit_df = 
    read_csv(
      "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", 
      na = c("NA", ".", ""), 
col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c"))

nyctransit_df =
  nyctransit_df |>
  janitor::clean_names() |>
  select(
    line, station_name, station_latitude, station_longitude,
    route1:route11, entry, vending, entrance_type, ada
  ) |>
  mutate(entry = case_match(entry, "YES" ~ TRUE, "NO" ~ FALSE))
      
```
This dataset consists of information on each entrance and exit of NYC subway situations. Variables included are the following:  station name, station latitude/longitude, routes served, subway line, whether the entrance allows entry, if there is vending, the entrance types and ADA compliance. 

To clean the data further I standardized the variable names to only have the variables of interest. I also converted entry from a character variable to a logical variable. After I ran the chunk I observed that the final dataset has `r nrow(nyctransit_df)` rows and `r ncol(nyctransit_df)` columns. Since route information is from route 1 through route 11 the route number is stored in the column names. 

```{r}
nyctransit_df |>
  distinct(line, station_name) |>
  nrow()
```
From the dataset we can see that, there are `r nrow(distinct(nyctransit_df, line, station_name))` distinct stations. 

```{r}
nyctransit_df |>
  filter(ada) |>
  distinct(line, station_name) |>
  nrow()
```

From the dataset we can see that there are `r nrow(distinct(filter(nyctransit_df, ada), line, station_name))` ADA compliant stations.

```{r}
nyctransit_df |>
  filter(vending == "NO") |>
  pull(entry) |>
  mean()
```
The proportion of entrances/exits without vending that allow entry is `r round(mean(filter(nyctransit_df, vending == "NO")$entry), 3)`.

```{r}
#Reformatting data so that route number and route name are distinct variables
nyctransit_tidy_df =nyctransit_df |>
  mutate(across(route8:route11, as.character)) |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name",
    names_prefix = "route",
    values_drop_na = TRUE
  )

# Distinct stations that serve the A train
nyctransit_tidy_df |>
  filter(route_name == "A") |>
  distinct(station_name, line) |>
  nrow()

# A train stations that are ADA compliant
nyctransit_tidy_df |>
  filter(route_name == "A", ada) |>
  distinct(station_name, line) |>
  nrow()
```

From the dataset provided we can see that, `r nyctransit_tidy_df |> filter(route_name == "A") |> distinct(station_name, line) |> nrow()` distinct stations serve the A train.

Of the stations that serve the A train, `r nyctransit_tidy_df |> filter(route_name == "A", ada) |> distinct(station_name, line) |> nrow()` are ADA compliant.

# Question 2

```{r}
#Read and Clean Mr. Trash Wheel sheet

library(readxl)
mrtrash_wheel_df =
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N742"
  ) |>
  janitor::clean_names() |> 
 drop_na(dumpster) |>
  mutate(sports_balls = as.integer(round(sports_balls)))
```
```{r}

#Combining with Mr. Trash Wheel Dataset to produce a single tidy dataset
library(tidyverse)
library(readxl)

mrtrash_wheel_df = 
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N742"
  ) |>
  janitor::clean_names() |> 
 drop_na(dumpster) |>
  mutate(
    sports_balls = as.integer(round(sports_balls)),
         year = as.integer (year),
         month = str_trim(month),
         trash_wheel = "Mr. Trash Wheel"
  )

proftrash_wheel_df =
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M139",
    col_types = c(
      "numeric", "text", "numeric", "date", "numeric", "numeric",
      "numeric", "numeric", "numeric", "numeric", "numeric", "numeric",
      "numeric"
    )
  ) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.integer(year),
    month = str_trim(month),
    trash_wheel = "Professor Trash Wheel"
  )

gwynndatr_df = 
  read_excel(
     "data/202610 Trash Wheel Collection Data.xlsx",
     sheet = "Gwynnda the Good Wheel of the W",
     range = "A2:L394"
  ) |> 
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(
    year = as.integer(year),
    month = str_trim(month),
    trash_wheel = "Gwynnda"
  )

trash_wheel_df = 
  bind_rows (mrtrash_wheel_df, proftrash_wheel_df, gwynndatr_df) |> 
  relocate (trash_wheel)

```
```{r}
 # the total weight of trash collected by Professor Trash Wheel
prof_totalweight =
  trash_wheel_df |>
  filter(trash_wheel == "Professor Trash Wheel") |>
  pull(weight_tons) |>
  sum(na.rm = TRUE)

#the total number of cigarette butts collected by Gwynnda in June of 2022
gwynnda_2022cigg =
  trash_wheel_df |>
  filter(trash_wheel == "Gwynnda", month == "June", year == 2022) |>
  pull(cigarette_butts) |>
  sum(na.rm = TRUE)

```
The final combined dataset, `trash_wheel_df`, contains `r nrow(trash_wheel_df)` observations, where each observation is a single dumpster collection by Mr. Trash Wheel, Professor Trash Wheel, or Gwynnda, with `r ncol(trash_wheel_df)` variables. The data span `r year(min(trash_wheel_df$date))` to `r year(max(trash_wheel_df$date))`. Key variables include `trash_wheel` (which wheel collected the trash), `dumpster` (the dumpster number), `month`, `year`, and `date`, along with `weight_tons` and `volume_cubic_yards`, and counts of specific items such as `plastic_bottles`, `cigarette_butts`, `wrappers`, and `homes_powered`. The variable `sports_balls` is available only for Mr. Trash Wheel, and `glass_bottles` is not recorded for Gwynnda, so those values are `NA` in the combined data. For available data, Professor Trash Wheel collected a total of `r round(prof_totalweight, 2)` tons of trash. Gwynnda collected a total of `r format(gwynnda_2022cigg, big.mark = ",")` cigarette butts in June of 2022.