---
title: "p8105_hw2_sm5975"
author: "Sania Sheifuddin"
date: "2026-10-07"
output: github_document
---

Read and clean the data; retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert the entry variable from character (YES vs NO) to a logical variable (case_match function may be useful).

Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far, and give the dimension (rows x columns) of the resulting dataset. Are these data tidy?

# Question 1 
```{r}
library(tidyverse)
nyctransit_df = 
    read_csv(
      "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv", 
      na = c("NA", ".", ""), 
col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c"))

nyctransit_df =
  nyctransit_df |>
  janitor::clean_names() |>
  select(
    line, station_name, station_latitude, station_longitude,
    route1:route11, entry, vending, entrance_type, ada
  ) |>
  mutate(entry = case_match(entry, "YES" ~ TRUE, "NO" ~ FALSE))
      
```
This dataset consists of information on each entrance and exit of NYC subway situations. Variables included are the following:  station name, station latitude/longitude, routes served, subway line, whether the entrance allows entry, if there is vending, the entrance types and ADA compliance. 

To clean the data further I standardized the variable names to only have the variables of interest. I also converted entry from a character variable to a logical variable. After I ran the chunk I observed that the final dataset has `r nrow(nyctransit_df)` rows and `r ncol(nyctransit_df)` columns. Since route information is from route 1 through route 11 the route number is stored in the column names. 

```{r}
nyctransit_df |>
  distinct(line, station_name) |>
  nrow()
```
From the dataset we can see that, there are `r nrow(distinct(nyctransit_df, line, station_name))` distinct stations. 

```{r}
nyctransit_df |>
  filter(ada) |>
  distinct(line, station_name) |>
  nrow()
```

From the dataset we can see that there are `r nrow(distinct(filter(nyctransit_df, ada), line, station_name))` ADA compliant stations.

```{r}
nyctransit_df |>
  filter(vending == "NO") |>
  pull(entry) |>
  mean()
```
The proportion of entrances/exits without vending that allow entry is `r round(mean(filter(nyctransit_df, vending == "NO")$entry), 3)`.

```{r}
#Reformatting data so that route number and route name are distinct variables
nyctransit_tidy_df =nyctransit_df |>
  mutate(across(route8:route11, as.character)) |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name",
    names_prefix = "route",
    values_drop_na = TRUE
  )

# Distinct stations that serve the A train
nyctransit_tidy_df |>
  filter(route_name == "A") |>
  distinct(station_name, line) |>
  nrow()

# A train stations that are ADA compliant
nyctransit_tidy_df |>
  filter(route_name == "A", ada) |>
  distinct(station_name, line) |>
  nrow()
```

From the dataset provided we can see that, `r nyctransit_tidy_df |> filter(route_name == "A") |> distinct(station_name, line) |> nrow()` distinct stations serve the A train.

Of the stations that serve the A train, `r nyctransit_tidy_df |> filter(route_name == "A", ada) |> distinct(station_name, line) |> nrow()` are ADA compliant.